Large-scale spatiotemporal variation in vital rates and population dynamics of an alpine bird

### 2024-09-12 Manuscript_Nater_etal.qmd
Added additional sections on post-hoc calculations

We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023]. For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance, see package vignette for specifics). We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible. We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility. Using NIMBLE's standard samplers, we then ran 5 MCMC chains of 150k iterations each. We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters). In addition to the main model run, we implemented a second version of the model that did not use telemetry data to assess the potential impacts of auxiliary data from a single location on parameter estimates.

## Post-hoc variance decomposition

@@

## Reproducible workflow with "targets"

Reproducibility and ease of repeating analyses was a key focus when developing the multi-area IDSM. To that end, we set up the workflow as a "targets pipeline", implemented through the R package "targets" [@landau2021]. The pipeline contains a variety of options for controlling modelling decisions in the workflow such as the year range of data to consider, the level of spatial aggregation (i.e. reporting district vs. survey locality), whether to model time variation in survival and/or effects of rodent occupancy, whether to run MCMC chains sequentially or in parallel, etc. A visual representation of the pipeline is also displayed in @fig-workflow and for more details on pipeline implementation and options, we refer the reader to the GitHub repository: <https://github.com/ErlendNilsen/OpenPop_Integrated_DistSamp>.
## Reproducible workflows with "R targets" and "Nix"

>>>>

We implemented our multi-area IDSM in a Bayesian framework using NIMBLE version 1.0.1 [@valpine2017] in R version 4.3.1 [@R2023]. For the likelihood for line transect observation distances we used a custom half-normal distribution developed by Michael Scroggie in the "nimbleDistance" package (https://github.com/scrogster/nimbleDistance, see package vignette for specifics). We used non-informative uniform priors for all parameters, but used biologically sensible boundaries where possible. We simulated complete sets of initial values for all model nodes prior to model running and using pre-defined seeds to ensure reproducibility. Using NIMBLE's standard samplers, we then ran 5 MCMC chains of 150k iterations each. We discarded the first 75k samples of each chain as burn-in, and thinned the remainder by a factor 25, resulting in a final joint posterior containing a total of 5 x 3k = 15k samples (note that high thinning rates were necessary to constrain memory load of the joint posterior, which included 314568 monitored parameters). In addition to the main model run, we implemented a second version of the model that did not use telemetry data to assess the potential impacts of auxiliary data from a single location on parameter estimates.
## Follow-up analyses

### Post-hoc variance decomposition

@@

### Calculation of additional demographic parameters

For presentation and interpretation of results, we calculated additional key demographic parameters from the model posteriors. First, we calculated area- and year-specific average population densities, $meanDens_{x, t}$, by averaging $D_{juv, x, j, t} + D_{ad, x, j, t}$ over all transect lines $j$ within area $x$ in year $t$. We then proceeded to derive area- and year-specific realized population growth rates as $meanDens_{x, t+1} / meanDens_{x, t}$. Additionally, we calculated generation time for each area using two different approaches: per-generation population growth rate [@caswell2000] and elasticity of asymptotic growth rate with respect to fecundity [@brooks2001].

### Parameter and sampling correlations

We also conducted correlation analyses on the posterior samples to a) check for potential evidence for vital rate trade-offs and/or density dependence and b) assess to what degree the former may be masked by sampling correlation. For a), we calculated Pearson's correlation coefficients between area-specific time series of estimated vital rate, population density, and population growth rate for each posterior sample. For b), we calculated Pearson's correlation coefficients between area- and time-dependent survival probabilities and recruitment rates across all posterior samples.

## Reproducible workflows with "R targets" and "Nix"
